Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-07-23 manuscripts/reply/reply-main.tex
More editing of reply ms.

\change{
The strong preference for the narrowest priors on divergence times used by
\citet{Hickerson2013} suggest their approach is biased toward models with less
parameter space and, as a consequence, will estimate model-averaged posteriors
dominated by models that exclude true values of the model's parameters.
}
We explored this possibility in two ways.
First, we re-analyzed the Philippines dataset using the model-averaging
approach of \citet{Hickerson2013}, but set one of the prior models with a
uniform prior on divergence times that is unrealistically narrow and almost
certainly excludes most, if not all, of the true divergence times of the 22
taxon pairs.
If small likelihoods of large models cause the method to prefer models with
less parameter space, we expect \msb will preferentially sample from this
incorrect model yielding a posterior that is incorrect (i.e., the
model-averaged posterior will be dominated by an incorrect model that excludes
the truth).
Second, we generated simulated datasets for which the divergence times are
drawn from an exponential distribution and applied the approach of
\citet{Hickerson2013} to each of them to see how often the method excludes the
truth.

@@

Furthermore, our results demonstrate that the approach of \citet{Hickerson2013}
consistently infers highly clustered divergences across all the \divt{max} we
simulated (Figure~\ref{figPower4}A--D \& S\ref{figPower6}A--F).
The method is most likely to infer the extreme case of a single divergence
event shared across all 22 taxa when populations diverged randomly over the
past \globalcoalunit generations.
When divergences are random over the past $8\globalpopsize$ generations, the
most likely inference is only two divergence events, and a single
divergence is still estimated in more than 10\% of the replicates.
It is interesting to note that as \divt{max} increases, but before the
estimates are finally pulled away from $\numt{} = 1$, the distribution of
\numt{} estimates closely mirror the U-shaped prior on divergence models
used by \msb (see Figure~\ref{figPower4}E and Oaks et al.'s
(\citeyear{Oaks2012}) Figure 5B).
This suggests this U-shaped prior coupled with the small marginal likelihoods
of models with many divergence parameters, is a major cause of the method's
bias toward clustered divergence models; this is consistent with
\citet{Oaks2012} and confirmed by \citet{Oaks2014dpp}.

Looking at our simulation results in terms of the posterior probability of the
dispersion index of divergence times supporting the extreme case of one
divergence event (i.e., $p(\vmratio{} < 0.01 \given \ssSpace)$), we find the
method strongly supports one divergence in greater than 27\% of the replicates
across all the \divt{max} we simulated (Figure~\ref{figPower4}E--H \&
S\ref{figPower6}G--L);
following \citet{Hickerson2013}, we use a Bayes factor of greater than 10 as
the criterion for incorrect inference of a single divergence event.
Furthermore, there is strong support for a single divergence event in
more than 90\% of the replicates when divergences are random over the past
$2.4\globalpopsize$ generations, and more than 60\% when over the past
$3.2\globalpopsize$ generations (Figure~\ref{figPower4}E--H \&
S\ref{figPower6}G--L).

Our results show that the empirical Bayesian model-averaging approach leads to
erroneous support for highly clustered divergences when populations diverge
randomly over the last $8\globalpopsize$ generations.
To put this on the scale roughly consistent with a vertebrate mitochondrial
locus, assuming a mutation rate of $1\e{-8}$ per site per generation, this
translates to 10 million generations.
Assuming a mutation rate consistent with nuclear loci of $1\e{-9}$, this is 100
million generations.
% Given the empirically Bayesian model-averaging approach of \citet{Hickerson2013}
% lacks power to detect random variation in divergence times over such
% timescales, it is difficult to justify the risk of the approach to exclude
% regions of parameter space containing the truth.

>>>>

\change{
The strong preference for the narrowest priors on divergence times used by
\citet{Hickerson2013} suggests their approach is biased toward models with less
parameter space and, as a consequence, will estimate model-averaged posteriors
dominated by models that exclude true values of the model's parameters.
}
We explored this possibility in two ways.
First, we re-analyzed the Philippines dataset using the model-averaging
approach of \citet{Hickerson2013}, but set one of the prior models with a
uniform prior on divergence times that is unrealistically narrow and almost
certainly excludes most, if not all, of the true divergence times of the 22
taxon pairs.
If small likelihoods of large models cause the method to prefer models with
less parameter space, we expect \msb will preferentially sample from this
incorrect model yielding a posterior that is incorrect (i.e., the
model-averaged posterior will be dominated by an incorrect model that excludes
the truth).
Second, we generated simulated datasets for which the divergence times are
drawn from an exponential distribution and applied the approach of
\citet{Hickerson2013} to each of them to see how often the method excludes the
truth.

@@

Furthermore, our results demonstrate that the approach of \citet{Hickerson2013}
consistently infers highly clustered divergences across all the \divt{max} we
simulated (Figure~\ref{figPower4}A--D \& S\ref{figPower6}A--F).
If we follow \citet{Hickerson2013} and use a Bayes factor of greater than 10 as
the criterion for incorrect inference of a single divergence event, we find
that the approach will often strongly support the extreme case of one
divergence event across all our simulation conditions (Figure~\ref{figPower4}E--H \&
S\ref{figPower6}G--L).
Overall, the empirical Bayesian model-averaging approach leads to erroneous
support for highly clustered divergences when populations diverge randomly over
the last $8\globalpopsize$ generations.
To put this on a timescale roughly consistent with a vertebrate mitochondrial
locus, assuming a mutation rate of $1\e{-8}$ per site per generation, this
translates to 10 million generations.
Assuming a mutation rate consistent with nuclear loci of $1\e{-9}$, this is 100
million generations.

% The method is most likely to infer the extreme case of a single divergence
% event shared across all 22 taxa when populations diverged randomly over the
% past \globalcoalunit generations.
% When divergences are random over the past $8\globalpopsize$ generations, the
% most likely inference is only two divergence events, and a single
% divergence is still estimated in more than 10\% of the replicates.
It is interesting to note that as \divt{max} increases, but before the
estimates are finally pulled away from $\numt{} = 1$, the distribution of
\numt{} estimates closely mirror the ``U-shaped'' prior on divergence models
used by \msb (see Figure~\ref{figPower4}E and Oaks et al.'s
(\citeyear{Oaks2012}) Figure 5B).
This suggests the U-shaped prior coupled with the small marginal likelihoods
of models with many divergence parameters is a major cause of the method's
bias toward clustered divergence models; this is consistent with
\citet{Oaks2012} and confirmed by \citet{Oaks2014dpp}.

% Looking at our simulation results in terms of the posterior probability of the
% dispersion index of divergence times supporting the extreme case of one
% divergence event (i.e., $p(\vmratio{} < 0.01 \given \ssSpace)$), we find the
% method strongly supports one divergence in greater than 27\% of the replicates
% across all the \divt{max} we simulated (Figure~\ref{figPower4}E--H \&
% S\ref{figPower6}G--L);
% following \citet{Hickerson2013}, we use a Bayes factor of greater than 10 as
% the criterion for incorrect inference of a single divergence event.
% Furthermore, there is strong support for a single divergence event in
% more than 90\% of the replicates when divergences are random over the past
% $2.4\globalpopsize$ generations, and more than 60\% when over the past
% $3.2\globalpopsize$ generations (Figure~\ref{figPower4}E--H \&
% S\ref{figPower6}G--L).

% Given the empirically Bayesian model-averaging approach of \citet{Hickerson2013}
% lacks power to detect random variation in divergence times over such
% timescales, it is difficult to justify the risk of the approach to exclude
% regions of parameter space containing the truth.
